Grove: A Convergent Collaborative Structure-Editor Calculus

### 2022-08-12 introduction.tex
still working on the intro

TODO: fix the following warning which are reported by LaTeX but not Overleaf: Class acmart Warning: A possible image without description on input line 321.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Programming is an increasingly collaborative activity.
Today, collaborative editing services are either provided directly by an asynchronous,
diff-and-merge-based version control system like git, darcs, or pijul,
or they are integrated into a real-time editor built on top of a distributed consistency protocol such as OT or CRDT.

Diff-and-merge is not an ideal collaborative editing solution.
Since diffs do not provide information on the original scope or ordering of edits,
the merging process must reconstruct an approximation, e.g., as a sequence of line additions and deletions.
Clearly, in the context of editing programs,
a line-based merge is not only less likely to succeed than one which retains finer grained information about changes over time,
it is also less likely to produce useful information about conflicting edit states, when they arise.

Operational Transformation (OT), on the other hand, goes perhaps too far in the opposite direction.
OT is a distributed system architecture that coordinates the proliferation and maintainance of edit states and actions.
Industrial-scale collaborative editors such as Google Docs tend to employ OT
as it provides strong theoretical guarantees on the eventual consistency of causally-ordered events
while remaining suitable for real-time applications.
It does so by maintaining a history of recent edits, or operations, while transforming individual edits to preserve a causal ordering.
But fully featured distributed systems that maintain causal orderings of arbitrary events are hard to design well,
so in practice OT-based systems tend to be restricted to linear text editing or have fewer features than comparable single-user alternatives.

Convergent Replicated Data Types (CRDTs) were proposed as a reaction to the complexity of Operational Transformation,
though the claim is currently under strutiny by the research community.
A CRDT is essentially a distributed data structure with an API for building and interacting with it.
CRDTs avoid the need for explicit state synchronization by requiring that allowable edit actions inhabit a lattice.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% start with something that everyone can accept: ``programming is collaborative.''

% briefly introduce 3 types: diff, OT, CRDT

% tease out some common problems (tree-native, collaborative)

% propose a solution

% general motivation: programming is a collaborative activity

% today usually takes the form of coarse-grained solutions, e.g, git

% if you try to use eg google docs for this, there are several problems
% (gap problem, diffs don't handle repositioning well)

% diff is not really an ideal way of communicating change: e.g., how repositioning gets mishandled

% CRDT: tree-native vs text. think about obvious limitation of existing tools

% algorithms for collaborative text editing are well developed
% natural fit for text
% not a natural fit for trees
% so if you want to build a grove, a more natural approach is needed

% a lot of collaborative editing of other sorts (not necessarily programming) requires a tree model,
% but that problem doesn't seem to have been study directly (maybe except JSON CRDT).

% A \emph{collaborative structure editor} is an editor that
% (1) allows multiple concurrent users to work on a shared document, while also
% (2) providing structure-aware editor services such as projectional editing, syntax highlighting, or automatic code folding.
% %
% Collaborative editing research focuses on the design and implementation of real-time, multi-user, character-based communication systems,
% whereas structure editors typically presume a more complex document schema and then focus on some other aspect of the user experience.
% In both settings, preservation of user intent is a core technical challenge.
% %
% Although collaborative editors and structure editors have overlapping goals (optimal user experience)
% and complementary design challenges (subject-subject versus subject-object harmony),
% to our knowledge, there is no comprehensive, principled account of their combined use.

% Since collaborative editors are essentially distributed systems, existing work tends to focus on extensions to distribution protocols.
% Lots of examples using OT. (Google Docs)
% OT is complex and largely textual.
% OT can make sense for real time systems: users typically change one character at a time, and instant feedback can help to prevent conflicts.
% On the other hand, OT system designs can be difficult to extend.

% Alternatively, there's CRDT. (Peritext?)
% CRDT is easier to implement, but harder to design for.
% There's an All-CRDT editor---it turned out to be not so realistic. (what's it called again?)

% Structure editing has been a recurring theme in the computer science literature since at least Engelbart's ``Mother of All Demos.''
% Provides automation for domain experts and reduces the barrier to entry for everyone else.
% Popular for editing programs, i.e., for programming language-specific editors.

% However, modern program editors typically disable editor services (like what?) when the document is not in a consistent state,
% a phenomenon called the ``gap problem.''
% Of course, in the presence of multiple concurrent users, the problem gets worse.

% In a collaborative setting,
% Hazel is a structure editor with support for advanced editing services, (e.g., semantic actions?).

\newpage
